---
output:
  pdf_document: default
  html_document: default
---


\newpage
```{r , include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	eval=TRUE,
	message = FALSE,
	warning = FALSE,
	comment = NA
)

```

# EM Algorithm

## Time Series plot describing dependence of Z and Y versus X.

```{r}
#1
library(ggplot2)

physical_data = read.csv("physical1.csv")
#head(physical_data)

ggplot(physical_data) + 
  geom_line(aes(X, Y, color='Y')) + 
  geom_line(aes(X, Z, color='Z')) + 
  ggtitle("Plot of Y and Z vs X")

```

Both Y and Z in this plot have a similar patter, just the scale of the oscillation is different. Z oscilates with an higher amplitude than Y. Both Y and Z decay over time in a similar manner.

## Derive an EM algorithm that estimates $\lambda$.

$$ 
Y_i \approx  exp(\frac{X_i}{\lambda}), \\
Z_i \approx exp(\frac{X_i}{2 * \lambda}) 
$$
Where $\lambda$ is an unknown parameters. The goal is to derive the EM algorithm that estimates $\lambda$.

$$
\begin{split}
L(\lambda|Y,Z) & = \prod_{i=1}^{n}f(Y) \times \prod_{i=1}^{n}f(Z) \\
 & = \prod_{i=1}^{n} \frac{X_i}{\lambda} \cdot e^{-\frac{X_i}{\lambda}Y_i} \times \prod_{i=1}^{n}\frac{X_i}{2\lambda} \cdot e^{-\frac{X_i}{\lambda}Z_i}\\
 & = \frac{X_1  \cdot \dotso \cdot X_n}{\lambda^n} \times e^{-\frac{1}{\lambda}\sum_{1}^{n}{X_i Y_i}} \times \frac{X_1  \cdot \dotso \cdot X_n}{(2\lambda)^n} \times e^{-\frac{1}{2\lambda}\sum_{1}^{n}{X_i Z_i}} \\
 lnL(\lambda|Y,Z) & = \sum_{i=1}^{n}ln(X_i)-nln(\lambda)-\frac{1}{\lambda}\sum_{i=1}^{n}{X_i Y_i}~+\sum_{i=1}^{n}ln(X_i)-nln(2\lambda)-\frac{1}{2\lambda}\sum_{i=1}^{n}{X_iZ_i}
\end{split}
$$

###E-step : Derive Q function

$$
\begin{split}
Q(\theta,\theta^k) & =E[~loglik(\lambda|Y,Z)~|~\lambda^k ,(Y,Z)] \\
&= \sum_{i=1}^{n}ln(X_i)-nln(\lambda)-\frac{1}{\lambda}\sum_{i=1}^{n}{X_i Y_i}~+\sum_{i=1}^{n}ln(X_i)-nln(2\lambda)\\
&~~~~~-\frac{1}{2\lambda}\bigg[\sum_{i=1}^{n}{X_iZ_i}~+m\cdot X_i\cdot\frac{2\lambda_{k-1}}{X_i}\bigg]
\end{split}
$$

Here, we are taking expectation on the missing values in Z, so we need to seperate the $Z_{obs}$ and $Z_{miss}$.  
Here we are assuming there are 'm' missing Z values. $\lambda_{k}$ is the lambda value from the previous iteration.


###M-step
We calculate the maximum likelihood estimate ($\lambda_{MLE}$) of the parameters by taking the derivative with respect to $\lambda$. Repeat till estimate converges. 

$$
\begin{split}
-\frac{n}{\lambda}-\frac{n}{\lambda}~+\frac{\sum_{i=1}^{n}{X_i Y_i}}{\lambda^2}~+\frac{\sum_{i}^{m}{X_iZ_i}~+m\cdot2\lambda_{k-1}}{2\lambda^2}&:=0\\
-2\lambda(2n)+2\sum_{i=1}^{n}{X_iY_i}~+\sum_{i=1}^{n}{X_iZ_i}+m\cdot2\lambda_{k-1} & := 0\\
\\
\lambda=\frac{\sum_{i=1}^{n}{X_iY_i}+\frac{1}{2}\sum_{i=1}^{n}{X_i}{Z_i}+m\cdot\lambda_{k-1}}{2n}
\end{split}
$$

## Implement EM Algorithm.

```{r}
##3
EM.Exp<-function(data, eps, kmax, lambda_0){
  X <- data$X; Y <- data$Y; Z <- data$Z
  Xobs <- X[!is.na(Z)]
  Zobs <- Z[!is.na(Z)]
  Zmiss <- Z[is.na(Z)]
  n <- length(X)
  m <- length(Zmiss)
  k <<- 0
  llvalprev <- 0
  llvalcurr <- lambda_0
  cat("Initial lambda value : ", llvalcurr, "\n")
  
  while ((abs(llvalprev-llvalcurr)>eps) && (k<(kmax+1))){
    llvalprev<-llvalcurr
    llvalprev <- llvalcurr
    llvalcurr <- (sum(X*Y)+sum(Xobs*Zobs)/2+m*llvalprev)/(2*n)
    k <<- k+1
  }
  return(c(llvalprev,llvalcurr,k))
}
lmb = EM.Exp(physical_data,0.001,100,100)
cat("The final converged lambda value : ", lmb[1], "\n")
cat("Number of steps taken to converge : ", lmb[3], "\n")
lambda <- lmb[2]
new_data <- physical_data
new_data$E_Y <- lambda/physical_data$X
new_data$E_Z <- 2*lambda/physical_data$X

```

We initialized $\lambda_{0}$ to 100 and it got converged at value $\lambda_{Converged}$ to 10.69587

We then went on to calculate the Expected values for Y and Z using this $\lambda$. Since Y and Z belong to Exponential distribution, the formula to compute the Expected value for them is : 
$$E[Y]=\frac{\lambda}{X_i}~,~~E[Z]=\frac{2\lambda}{X_i}$$

```{r}
#4
ggplot(data=new_data) +
  geom_line(aes(x = X, y = Y, colour = "Y")) +
  geom_line(aes(x = X, y = Z, colour = "Z")) +
  geom_line(aes(x = X, y = E_Y, colour = "E_Y")) +
  geom_line(aes(x = X, y = E_Z, colour = "E_Z")) +
  ggtitle("Expected value of Y and Z")

```

From the plot we can see that the E[Y] and E[Z] capture the trend of Y and Z respectively, so we can conclude that the estimated lambda using EM algorithm is a good estimate.