---
title: "Group_18_Lab2"
author: "Sridhar, Omkar"
date: "September 20, 2018"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(plotly)
library(xlsx)
library(MASS)
library(gridExtra)
```


##Assignment 1 

#### 1
In our first plot, linoleic is a continous variable with colored in hue's of blue. The channel capacity for 
hue is only 10 levels and the problem of occusion with the data also deteriorates our understanding of the plot.
In the second plot, We have linoleic variable segmented into 4 groups, this gives us a quicker understanding of 
the data showing us the relative values for the groups. The perception problem of relative judgement is affected 
as color hue comes with the highest error in human beings.

```{r}
ol = read.csv("olive.csv", header = T, row.names = 1)
head(ol, 10)

ggplot(ol, aes(palmitic, oleic, col = linolenic)) + geom_point()

disc = cut_interval(ol$linolenic, 4)
ggplot(ol, aes(palmitic, oleic, col = disc)) + geom_point()
```

####2

The 2nd plot is the easiest to analyse the plot with linolenic segmented into 4 groups. The size mapping creates
the problem of occlusion due to overlapping. The orientaion angle map does not help either as the scatter plot 
many observations creates a high relative judement error.
a)With Color hue 10 levels of feature can be percieved and 3.1bits can be decoded,With Color Brightness 5 levels 
and 2.1bits can be decoded.
b)With size of object 4-5levels of feature can be percieved depending on human subject's 
individualistic abilities, and 2.2bits can be decoded for this aesthetic.
c)line orientation : 3bits can be decoded for this feature.

```{r}
ggplot(ol, aes(palmitic, oleic, col = disc)) + geom_point()
ggplot(ol, aes(palmitic, oleic, size = disc)) + geom_point()
ggplot(ol, aes(palmitic, oleic)) + geom_point() + 
    geom_spoke(angle = ol$linolenic, radius = 40)
```

####3

Treisman's theory of preattentive processing is showcased in this example. With no segmentation done by Region,
we do not easily identify the boundary, but with the second plot we see the same much quickly due to preattentive
preprocessing of contrast and luminance.

```{r}
ggplot(ol, aes(oleic, eicosenoic, col = Region)) + geom_point()
ggplot(ol, aes(oleic, eicosenoic, col = cut_interval(Region,3))) + geom_point()
```

####4

The 3 colors are each mapped with contrast and size and these feature maps are parallely processed in our brain
as this creates a problem of preattentive preprocessing while analysing 27 different types of observation as
we tend to wrongly group the data in our perception

```{r}
ggplot(ol, aes(ol$oleic, ol$eicosenoic, col = cut_interval(ol$linoleic, 3),
              shape = cut_interval(ol$palmitic, 3), 
              size = cut_interval(ol$palmitoleic, 3))) + geom_point()
```

####5

Size, contrast and shape are individual feature maps that are linked to different colors and hence preattentive 
preprocessing helps in this case.

```{r}
ggplot(ol, aes(ol$oleic, ol$eicosenoic, col = ol$Region,
              shape = cut_interval(ol$palmitic, 3), 
              size = cut_interval(ol$palmitoleic, 3))) + geom_point()
```

####6

Relative Judgement due to area is very high due to the plot made as a pie size as the dominant group of South-
Apulia looks much larger than the other groups.

```{r}
p <- plot_ly(ol, labels = ~Area, type = 'pie', showlegend = FALSE) %>%
  layout(title = 'Pie Chart Area')
p
```

####7
It is hard to look for outliers in the contour plot compared to the scatter plot. In the contour plot it shows we have 5 peak values but you wont be able to spot any difference for it in the scatter plot. The extreme values are not plotted in the contour plot. It is also hard to figure out clusters in the contour plot compared to a scatter plot which is a big issue in this plot.

```{r}
ggplot(ol, aes(ol$linoleic, ol$eicosenoic)) + geom_density2d()
ggplot(ol, aes(ol$linoleic, ol$eicosenoic)) + geom_point()
```


##Assignment 2 

####1 - Read Data

The columns vary a lot in the range. Some values like BAvg are averages so are in a range of 0.235 to 0.282, while values like TB are in the range 2090 to 2615. This is the reason scaling is required before we apply Non- metric MDS. Scaling the data gets all the values in the same range, this would allow the NMDS algorithm to reduce the dimensions of the data more efficiently.
```{r echo=FALSE}
bball = read.xlsx("baseball-2016.xlsx", sheetName = "Sheet1", header = TRUE,
                  row.names = 1)
head(bball)
```


####2 - Non-metric MDS

It is hard to see a difference between the legues in this plot. We could say that the National League(NL) teams are spread out away from the origin, and the Anerican League(AL) teams are more centered towards the origin. 

The points are well spread out so it is hard to tell if a MDS component is providing better differentiation between the leagues. In my opinion V1 was doing a better split between the leagues compared to V2. 

According to this plot "Boston Red Sox" and "Atlanta Braves" look like outliers. 

```{r warning=FALSE, echo=FALSE}
bball.numeric = as.data.frame(scale(bball[,2:27]))
distance = dist(bball.numeric)
res = isoMDS(distance, k=2, p=2)
coords = res$points

coordsMDS = as.data.frame(coords)
coordsMDS$name = rownames(coordsMDS)
coordsMDS$league = bball$League
plot_ly(coordsMDS, x=~V1, y=~V2, type="scatter", mode = "markers"
        , hovertext=~name, color= ~league)
```


####3 - Shepard Plot

MDS was able to decrease the stress value upto 15.6%. Given that the dataset had 26 dimension and getting it down to 2 dimensions, with stress level of 15.6 is good. 

Some of the observation pairs that were hard for MDS to map were - 

"Orkland Athletics and Milwaukee Brewers",
"NY Mets and Minnesota Twins",
"Minnesota Twins and Arizona Diamondbacks",
"Orkland Athletics and Chicago cubs",
"Pittsburg pirates and Chicago cubs"

```{r echo=FALSE}
sh <- Shepard(distance, coords)
delta <-as.numeric(distance)
D<- as.numeric(dist(coords))

n=nrow(coords)
index=matrix(1:n, nrow=n, ncol=n)
index1=as.numeric(index[lower.tri(index)])

n=nrow(coords)
index=matrix(1:n, nrow=n, ncol=n, byrow = T)
index2=as.numeric(index[lower.tri(index)])

plot_ly()%>%
  add_markers(x=~delta, y=~D, hoverinfo = 'text',
              text = ~paste('Obj1: ', rownames(bball)[index1],
                            '<br> Obj 2: ', rownames(bball)[index2]))%>%
  add_lines(x=~sh$x, y=~sh$yf)
```


####4 - 

Since V1 was spliting the leagues better, I plotted all the variables against it and found that "RBI" and "OPS" had a strong negative connection with V1. On searching for these on google we found, these two turned out to be really important factors in baseball to differentiate teams and rank them.

RBI(Runs batted in) - RBI is a statistic in baseball that credits a batter for making a play that allows a run to be scored. The top teams in the league have a high RBI. It is an important batting statistic in baseball.

OPS(On-base plus slugging) - OPS is a statistic calculated as a sum of players ability to get on base and hit with power. Usually the league leaders have players with highest OPS. This is an important factor that increases runs scored for a team.

Both of these RBI and OPS are important batting statistics in baseball.

```{r include=FALSE, echo=FALSE}
cols_bball = colnames(bball)
plots_bball = list()
for(i in 2:27){
  pl_name = paste("P", i, sep = '')
  col_name = cols_bball[i]
  plots_bball[[pl_name]] = ggplot(bball.numeric, aes_string("coordsMDS$V1", col_name)) + 
    geom_point() + geom_line(aes(x=0))
}
grid.arrange(grobs = plots_bball, ncol = 6, nrow = 6)

plots_bball$P2
plots_bball$P3
plots_bball$P4 #Runs per game
plots_bball$P5
plots_bball$P6 #AB
plots_bball$P7 #Runs **
plots_bball$P8 #Hits **
plots_bball$P9
plots_bball$P10
plots_bball$P11
plots_bball$P12 #RBI **
plots_bball$P13
plots_bball$P14
plots_bball$P15
plots_bball$P16
plots_bball$P17 #BAvg
plots_bball$P18
plots_bball$P19 #SLG
plots_bball$P20 #OPS  **
plots_bball$P21 #TB  **
plots_bball$P22
plots_bball$P23
plots_bball$P24
plots_bball$P25
plots_bball$P26
plots_bball$P27

plots_bball2 = list()
for(i in 2:27){
  pl_name = paste("P", i, sep = '')
  col_name = cols_bball[i]
  plots_bball2[[pl_name]] = ggplot(bball, aes_string("coordsMDS$V2", col_name)) + 
    geom_point() + geom_line(aes(x=0))
}
grid.arrange(grobs = plots_bball2, ncol = 6, nrow = 6)
plots_bball2$P2
plots_bball2$P3
plots_bball2$P4
plots_bball2$P5
plots_bball2$P6
plots_bball2$P7
plots_bball2$P8
plots_bball2$P9
plots_bball2$P10
plots_bball2$P11
plots_bball2$P12
plots_bball2$P13
plots_bball2$P14
plots_bball2$P15
plots_bball2$P16
plots_bball2$P17
plots_bball2$P18
plots_bball2$P19
plots_bball2$P20
plots_bball2$P21
plots_bball2$P22
plots_bball2$P23
plots_bball2$P24
plots_bball2$P25
plots_bball2$P26
plots_bball2$P27
```
```{r}
plots_bball$P12 #TB  **
plots_bball$P20 #OPS  **
```

#Appendix
```{r eval=FALSE}
library(ggplot2)
library(plotly)
library(xlsx)
library(MASS)
library(gridExtra)
#Assignment 1
#Q1
ol = read.csv("olive.csv", header = T, row.names = 1)
head(ol, 10)

ggplot(ol, aes(palmitic, oleic, col = linolenic)) + geom_point()

disc = cut_interval(ol$linolenic, 4)
ggplot(ol, aes(palmitic, oleic, col = disc)) + geom_point()


#Q2
ggplot(ol, aes(palmitic, oleic, col = disc)) + geom_point()
ggplot(ol, aes(palmitic, oleic, size = disc)) + geom_point()
ggplot(ol, aes(palmitic, oleic)) + geom_point() + 
    geom_spoke(angle = ol$linolenic, radius = 40)

#Q3
ggplot(ol, aes(oleic, eicosenoic, col = Region)) + geom_point()


#Q4
ggplot(ol, aes(ol$oleic, ol$eicosenoic, col = cut_interval(ol$linoleic, 3),
              shape = cut_interval(ol$palmitic, 3), 
              size = cut_interval(ol$palmitoleic, 3))) + geom_point()


#Q5
ggplot(ol, aes(ol$oleic, ol$eicosenoic, col = ol$Region,
              shape = cut_interval(ol$palmitic, 3), 
              size = cut_interval(ol$palmitoleic, 3))) + geom_point()

#Q6
p <- plot_ly(ol, labels = ~Area, type = 'pie', showlegend = FALSE) %>%
  layout(title = 'Pie Chart Area')
p

#Q7
ggplot(ol, aes(ol$linoleic, ol$eicosenoic)) + geom_density2d()
ggplot(ol, aes(ol$linoleic, ol$eicosenoic)) + geom_point()


#Assignment 2
#Q1

bball = read.xlsx("baseball-2016.xlsx", sheetName = "Sheet1", header = TRUE,
                  row.names = 1)
head(bball)

#Q2
bball.numeric = bball[,3:27]
distance = dist(bball.numeric)
res = isoMDS(distance, k=2, p=2)
coords = res$points

coordsMDS = as.data.frame(coords)
coordsMDS$name = rownames(coordsMDS)
coordsMDS$league = bball$League
plot_ly(coordsMDS, x=~V1, y=~V2, type="scatter", mode = "markers"
        , hovertext=~name, color= ~league)


#Q3
sh <- Shepard(distance, coords)
delta <-as.numeric(distance)
D<- as.numeric(dist(coords))

n=nrow(coords)
index=matrix(1:n, nrow=n, ncol=n)
index1=as.numeric(index[lower.tri(index)])

n=nrow(coords)
index=matrix(1:n, nrow=n, ncol=n, byrow = T)
index2=as.numeric(index[lower.tri(index)])

plot_ly()%>%
  add_markers(x=~delta, y=~D, hoverinfo = 'text',
              text = ~paste('Obj1: ', rownames(bball)[index1],
                            '<br> Obj 2: ', rownames(bball)[index2]))%>%
  add_lines(x=~sh$x, y=~sh$yf)

#Q4
bball$V1 = coordsMDS$V1
bball$V2 = coordsMDS$V2

cols_bball = colnames(bball)
dim(bball)[2]
plots_bball = list()
for(i in 2:27){
  pl_name = paste("P", i, sep = '')
  col_name = cols_bball[i]
  plots_bball[[pl_name]] = ggplot(bball, aes_string("V1", col_name)) + 
    geom_point() + geom_line(aes(x=0))
}
grid.arrange(grobs = plots_bball, ncol = 6, nrow = 6)

plots_bball$P2
plots_bball$P3
plots_bball$P4 #Runs per game
plots_bball$P5
plots_bball$P6 #AB
plots_bball$P7 #Runs **
plots_bball$P8 #Hits **
plots_bball$P9
plots_bball$P10
plots_bball$P11
plots_bball$P12 #RBI **
plots_bball$P13
plots_bball$P14
plots_bball$P15
plots_bball$P16
plots_bball$P17 #BAvg
plots_bball$P18
plots_bball$P19 #SLG
plots_bball$P20 #OPS  **
plots_bball$P21 #TB  **
plots_bball$P22
plots_bball$P23
plots_bball$P24
plots_bball$P25
plots_bball$P26
plots_bball$P27

plots_bball2 = list()
for(i in 2:27){
  pl_name = paste("P", i, sep = '')
  col_name = cols_bball[i]
  plots_bball2[[pl_name]] = ggplot(bball, aes_string("V2", col_name)) + 
    geom_point() + geom_line(aes(x=0))
}
grid.arrange(grobs = plots_bball2, ncol = 6, nrow = 6)
plots_bball2$P2
plots_bball2$P3
plots_bball2$P4
plots_bball2$P5
plots_bball2$P6
plots_bball2$P7
plots_bball2$P8
plots_bball2$P9
plots_bball2$P10
plots_bball2$P11
plots_bball2$P12
plots_bball2$P13
plots_bball2$P14
plots_bball2$P15
plots_bball2$P16
plots_bball2$P17
plots_bball2$P18
plots_bball2$P19
plots_bball2$P20
plots_bball2$P21
plots_bball2$P22
plots_bball2$P23
plots_bball2$P24
plots_bball2$P25
plots_bball2$P26
plots_bball2$P27
```


```{r}
```
