---
title: "Lab6_Group18"
author: "Sridhar, Omkar"
date: "17 oktober 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libs, message=FALSE, warning=FALSE}
library(visNetwork)
library(tidyverse)
library(igraph)
library(plotly)
library(seriation)
library(tourr)
```

#Assignment 1

The Bombing group and the Non Bombing groups are seperated by colors Red and Blue respectively. There are more people in the non bombing group. The clusters found in each of the groups are as follows -

- In the bombing group the two main clusters found are originated at "Jamal Zougam" and "Mohamed Chaoui".There are also some other smaller clusters .

- In the Non bombing group the main cluster found was originated at "Imad Eddin Barakat". There are many other smaller clusters in the Non bombing group. From the plot it looks like Imad had direct links to the bombing group.

###1. Network Plot Degree 0

```{r q1_1,message=FALSE, warning=FALSE}
nodes<-read.table("trainMeta.dat")
colnames(nodes)<-c("label","group")
nodes$id<-rownames(nodes)
nodes<-nodes[,c(3,1,2)]
nodes$title<-nodes$label
nodes$color<-ifelse(nodes$group==1,"red","blue")
nodes<-data.frame(nodes)

links<-read.table("trainData.dat")
colnames(links)<-c("from","to","value")
links<-data.frame(links)

weight_nodes<-graph.data.frame(d=links,vertices=nodes,directed = F)
degree_nodes<-degree(weight_nodes,mode="all")
nodes$value<-degree_nodes[match(nodes$id,names(degree_nodes))]


q1<-visNetwork(nodes,links)%>%visPhysics(solver="repulsion") %>%
  visOptions(highlightNearest = list(enabled = T,degree=0,
                                     hover = T),nodesIdSelection=TRUE,
             selectedBy = "group") 
q1
```



###2. Network Plot Degree 1

"Jamal Zougam" (born 1973 in Tangier) was one of six men implicated in the 2004 Madrid train bombings. He was detained on 13 March 2004, accused of multiple counts of murder, attempted murder, stealing a vehicle, belonging to a terrorist organisation and four counts of carrying out terrorist acts. Spain's El País newspaper reported that three witnesses testified to seeing him leave a rucksack aboard one of the bombed trains.

"Jamal Zougam" is the person with most number of connections in the network so he is the one who could spread information fastest.

```{r q1_2, message=FALSE, warning=FALSE}
q2<-visNetwork(nodes,links)%>%visPhysics(solver="repulsion") %>%
  visOptions(highlightNearest = list(enabled = T,degree=1,
                                     hover = T),nodesIdSelection=TRUE,
             selectedBy = "group")
q2
```

###3. Clusters by optimizing links

Yes I think the cluster I found centered by "Jamal Zougam" is the one most easily spotted in this plot also. This helps us identify clusters easily.

```{r q1_3, message=FALSE, warning=FALSE}
nodes1<-nodes
ceb<-cluster_edge_betweenness(weight_nodes)
nodes1$group<-ceb$membership
visNetwork(nodes1,links)%>%visIgraphLayout(layout = "layout_nicely")%>%
  visOptions(highlightNearest = list(enabled = T,degree=1,
             hover = T),nodesIdSelection=TRUE, selectedBy = "group") 
```


###4. Heatmap for finding clusters

Yes the main clusters identified using the reordered heatmap are the same spotted in the step 1 and 3. The two main clusters prominant in this plot are centered by "Jamal Zougam" and "Mohamed Chaoui". They had most number of connections with the Bombers and the Non bombers.

```{r q1_4, message=FALSE, warning=FALSE}
netm<-get.adjacency(weight_nodes,sparse = F)
colnames(netm)<-nodes$label
rownames(netm)<-nodes$label
rowdist<-dist(netm)

row_order<-seriate(rowdist,"HC")
order1<-get_order(row_order)
netm_reord<-netm[order1,order1]

plot_ly(z=~netm_reord,x=~colnames(netm_reord),
        y=~rownames(netm_reord),type="heatmap")%>% 
  layout(title = " Madrid Bombing Heatmap for finding  clusters")
```



##Assignment 2

###1

China and US were the largest producers of Oil and Coal throughout. China had a sudden rise in production of Coal from 2000 onwards. US was the highest producer of Oil throughout. 

```{r q2_1, message=FALSE, warning=FALSE}
oil_data = read.csv("Oilcoal.csv", header = TRUE, sep = ";", dec = ",")
oil_data$X = NULL
head(oil_data)


p1 <- oil_data %>%
  plot_ly(
    x = ~Oil, 
    y = ~Coal,
    color = ~Country,
    frame = ~Year, 
    type = 'scatter',
    mode = 'markers',
    text = ~Country, 
    hoverinfo = "text",
    size = ~Marker.size
  ) %>% 
  layout(
    xaxis = list(
      title = "Oil",
      zeroline = F
    ),
    yaxis = list(
      title = "Coal",
      zeroline = F
    )
  ) %>% 
  animation_opts(
    redraw = FALSE
  )
p1
```

###2

I found that France and Germany had quiet similar motion patterns. Though germany was higher in Coal and Oil compared to France throughout but they were moving up and down together for the whole time span. There was a sudden fall in Oil in both the countries between 1980 - 1983. Both the countries had a large drop. 
The reason for the drop was the Oil Price collapse in that period. That is the reason all the countries were showing this sudden drop.

```{r q2_2, message=FALSE, warning=FALSE}
fil_d = oil_data[oil_data$Country == "France"| oil_data$Country == "Germany",]
p2 <- fil_d %>%
  plot_ly(
    x = ~Oil, 
    y = ~Coal,
    color = ~Country,
    frame = ~Year, 
    type = 'scatter',
    mode = 'markers',
    text = ~Country, 
    hoverinfo = "text",
    size = ~Marker.size
  ) %>% 
  layout(
    xaxis = list(
      title = "Oil",
      zeroline = F
    ),
    yaxis = list(
      title = "Coal",
      zeroline = F
    )
  ) %>% 
  animation_opts(
    redraw = FALSE
  )
p2
```

###3

It is easier to look for values for a specific time in a barplot compared to the bubble plot. This is the main advantage bar plot has over bubble plot. Comparing between countries is also easier in a bar plot. 

The disadvantage of a bar plot over bubble plot is that we can analyse just a dimension of data in this. In bubble plot we can see how the points move compared to two dimensions of data, which is not possible in bar plot.

```{r q2_3, message=FALSE, warning=FALSE}
oil_data$oil_p = (oil_data$Oil/(oil_data$Oil+oil_data$Coal))*100
temp = oil_data[,]
temp$oil_p = 0
new_oil_data = rbind(oil_data, temp)

p3 <- new_oil_data %>%
  plot_ly(
    x = ~oil_p, 
    y = ~Country,
    split = ~Country,
    frame = ~Year, 
    type = 'scatter',
    mode = 'lines', 
    line = list(simplyfy = F, width = 20)
  ) %>% 
  layout(
    xaxis = list(
      title = "Oil_P",
      zeroline = F
    ),
    yaxis = list(
      title = "Country",
      zeroline = F
    )
  ) %>% 
  animation_opts(
    redraw = FALSE
  )
p3
```

###4

The elastic transition between values makes the movements more quick and precise as compared to the continuous movement. I found this to give a better understanding for the data movement. It had a pause in between the transitions which made it look better. These were advantages of using Elastic.

I did not find any disadvantages in this. I found it better than the continuous movement.

```{r q2_4, message=FALSE, warning=FALSE}
p4 <- new_oil_data %>%
  plot_ly(
    x = ~oil_p, 
    y = ~Country,
    split = ~Country,
    frame = ~Year, 
    type = 'scatter',
    mode = 'lines', 
    line = list(simplyfy = F, width = 20)
  ) %>% 
  layout(
    xaxis = list(
      title = "Oil_P",
      zeroline = F
    ),
    yaxis = list(
      title = "Country",
      zeroline = F
    )
  ) %>% 
  animation_opts(
    redraw = FALSE,
    easing = "elastic"
  )
p4
```

###5

I found that the projection near the value 6.8 on the scale best defined the clusters. I was able to spot two clusters using this projection.

Yes the clusters currospond to different year ranges. The first cluster has year ranges from 1965 to 1980, and the second cluster has the range 1980 to 2007. 


```{r q2_5, message=FALSE, warning=FALSE}
mat <- read.csv2("Oilcoal.csv",sep=";")
mat<-mat[,1:3]
mat <- mat %>%spread(Country, Coal)
mat_scaled <- rescale(mat[,2:9])

rownames(mat_scaled) <- mat[,1]

set.seed(12345)
#tour <- new_tour(mat, grand_tour(), NULL)
tour<- new_tour(mat_scaled, guided_tour(cmass), NULL)

steps <- c(0, rep(1/15, 200))
Projs<-lapply(steps, function(step_size){ 
  step <- tour(step_size)
  if(is.null(step)) {
    .GlobalEnv$tour<- new_tour(mat_scaled, guided_tour(cmass), NULL)
    step <- tour(step_size)
  }
  step
}
)

# projection of each observation
tour_dat <- function(i) {
  step <- Projs[[i]]
  proj <- center(mat_scaled %*% step$proj)
  data.frame(x = proj[,1], y = proj[,2], state = rownames(mat_scaled))
}

# projection of each variable's axis
proj_dat <- function(i) {
  step <- Projs[[i]]
  data.frame(
    x = step$proj[,1], y = step$proj[,2], variable = colnames(mat_scaled)
  )
}

stepz <- cumsum(steps)

# tidy version of tour data

tour_dats <- lapply(1:length(steps), tour_dat)
tour_datz <- Map(function(x, y) cbind(x, step = y), tour_dats, stepz)
tour_dat <- dplyr::bind_rows(tour_datz)

# tidy version of tour projection data
proj_dats <- lapply(1:length(steps), proj_dat)
proj_datz <- Map(function(x, y) cbind(x, step = y), proj_dats, stepz)
proj_dat <- dplyr::bind_rows(proj_datz)

ax <- list(
  title = "", showticklabels = FALSE,
  zeroline = FALSE, showgrid = FALSE,
  range = c(-1.1, 1.1)
)

# for nicely formatted slider labels
options(digits = 3)
tour_dat <- highlight_key(tour_dat, ~state, group = "A")
tour <- proj_dat %>%
  plot_ly(x = ~x, y = ~y, frame = ~step, color = I("black")) %>%
  add_segments(xend = 0, yend = 0, color = I("gray80")) %>%
  add_text(text = ~variable) %>%
  add_markers(data = tour_dat, text = ~state, ids = ~state, hoverinfo = "text") %>%
  layout(xaxis = ax, yaxis = ax,title="Animated guided tour of Coal consumption per Country")
tour
```

### Apendix
```{r, ref.label=knitr::all_labels(),echo=TRUE,eval=FALSE}
```