---
title: "Lab1"
author: "Sridhar Adhikarla"
date: "January 28, 2019"
output: 
        pdf_document :
            toc  : true
            toc_depth: 5
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE,
	comment= NA
)
```

# Question 1

## 1 Snippet 1 -

```{r}
#Q1
x1 <- 1/3
x2 <- 1/4
if(x1-x2 == 1/12){
  print("Subtraction is correct")
}else{
  print("Subtraction is wrong")
}
```

The way 1/3 is represented in r is makes the difference in this case. Since the representation 
of 1/3 is different, the subtraction (1/3 - 1/4) results in a different answer than 1/12.
The difference occurs in the 17th decimal place. The difference is very small but this is what
results in the inequality in the first case.

## 2 Snippet 2 -

```{r}
x1 <- 1
x2 <- 1/2
if(x1-x2 == 1/2){
  print("Subtraction is correct")
}else{
  print("Subtraction is wrong")
} 
```

1/2 is not such a complex fraction so the representation is easy for it and results in 
x2 being equal to (0.5). This is the reason  the equality is true in this case.
1 - 1/2 is 0.5 and 1/2 is also 0.5, resulting in the equality.

## 3 Improvement -

```{r}
x1 <- 1/3
x2 <- 1/4
a = formatC(x1-x2, digits = 10, format = 'f')
b = formatC(1/12, digits = 10, format = 'f')
if(a == b){
  print("Subtraction is correct")
}else{
  print("Subtraction is wrong")
} 
```

The way we can improve this is comparing only a few of the decimal places, like the first
10 decimal places of a number.

# Question 2

## 1

```{r}

#Q2
f =function(x){
  return(x)
}
derivative = function(f, x){
  eps = 10^-15
  d = (f(x+eps) - f(x))/(eps)
  return(d)
}
```

We created a function f that is an identity function, and we also created a derivative function that accepts as input a function that you want to find the derivative of and the value of x at which you want the derivative of the function and returns the derivative  value.

The derivative function is an approximation of the actual derivative, and for the function f the actual derivative for any value of x is equal to 1. We shold get an answer close to 1.

## 2

```{r}
derivative(f, 1)
```

For x=1 we get 1.110223, which is close to 1. 

## 3

```{r}
derivative(f, 100000)
```

For x=100000 the derivative turns out to be zero. Since the function we are finding the derivative for is an identity function, the numerator for the derivative turns out to be {epsilon(which is a very small number) + x - x}. This works fine for small values of x but as the values of x become larger the numerator becoms zero. This is the reason we get 0 for x = 100000. 

Since epsilon is a very small decimal the R starts ignoring it as the values of x become larger. So it gives us zero.


# Question 3

```{r}
#Q3
myVar = function(x){
  mu = (sum(x)^2)/length(x)
  s = sum(x^2)
  vr = (s - mu)/(length(x)-1)
  return(vr)
}
x = rnorm(10000, 10^8, 1)
Yi = matrix(0, nrow = 1, ncol = 10000)
for(i in 2:10000){
  Xi = x[1:i]
  Yi[i] = myVar(Xi) - var(Xi)
}
plot(1:10000, Yi)


myVar2 = function(x){
  mu = (sum(x)^2)/length(x)
  s = (x^2) - mu
  vr = (sum(s))/(length(x)-1)
  return(vr)
}
x = rnorm(10000, 10^8, 1)
Yi = matrix(0, nrow = 1, ncol = 10000)
for(i in 2:10000){
  Xi = x[1:i]
  Yi[i] = myVar2(Xi) - var(Xi)
}
plot(1:10000, Yi)
```


# Question 4

### Unscaled Matrix

```{r}
#Q4
data = read.csv("tecator.csv", row.names = 1)

y_data = as.matrix(data$Protein)
x_data = as.matrix(data[,- c(102)])

A = t(x_data)%*%x_data
b = t(x_data)%*%y_data

#solve(A, b)

cond_mat = kappa(A)
cond_mat
log2(cond_mat)
```

### Scaled Matrix

```{r}
#scaling data
x_data_scaled = scale(x_data)

A = t(x_data_scaled)%*%x_data_scaled
b = t(x_data_scaled)%*%y_data

#solve(A, b)

cond_mat_scaled = kappa(A)
cond_mat_scaled
log2(cond_mat_scaled)
```



# Appendix

```{r}
```