---
title: "Lab1_Sridhar"
author: "Sridhar Adhikarla (sriad858)"
date: "November 18, 2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE
  )
```

```{r message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
library(stargazer)
```


# Question 1

## a - Describe the 7 variables with mean values, standard deviations e.t.c.

```{r}
data_T1 = read.csv("T1-9.dat", sep = "\t", header = F)
colnames(data_T1) = c("Country", "100m", "200m", "400", "800", "1500", "3000", "marathon")

data_T1 %>% 
  head(5)

data_T1 %>%
  select(-Country) %>%
  stargazer(type = "text", 
    summary.stat = c("min", "p25", "median", "p75", "max", "median", "sd")
)
```

## b - Illustrate the variables with different graphs (explore what plotting possibilities R has)

```{r}
```

# Question 2

## a - Compute the covariance and correlation matrices for the 7 variables. Is there any apparent structure in them? Save these matrices for future use.

```{r}
```

## b - Generate and study the scatterplots between each pair of variables. Any extreme values?

```{r}
```

## c - Explore what other plotting possibilities R orders for multivariate data. Present other (at least two) graphs that you and interesting with respect to this data set.

```{r}
```

# Question 3

```{r}
```