---
title: "lab2_sridhar"
author: "Sridhar Adhikarla (sriad858)"
date: "September 28, 2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
#Required libs
library(astsa)
library(kernlab)
library(ggplot2)

set.seed(12345)
```

# Assignment 1

## a - Simualte values from AR(3)

The pcaf

```{r}
ar3 <- arima.sim(model = list(ar = c(0.8,-0.2,0.1)), n = 1000)

simulationCorrelation = function(ar3){
  df = ts.intersect("xt" = ar3, "xt1" = lag(ar3, k = -1), "xt2" = lag(ar3, k = -2))
  mod = lm(xt ~ xt1 + xt2, data = df)
  coeffs = c(mod$coefficients[2], mod$coefficients[3])
  xthPrime = lag(ar3, k = -3) - (coeffs[2]*lag(ar3, k = -1)+coeffs[1]*lag(ar3, k = -2))
  xtPrime = ar3 - (coeffs[1]*lag(ar3, k = -1)+coeffs[2]*lag(ar3, k = -2))
  
  return(cor(xthPrime, xtPrime))
}

par(mfrow=c(1,1),oma = c(0, 0, 2, 0))
pacfVal <- pacf(ar3, main=NA)$acf[3]
mtext("Practical ACF for AR(3) process", outer = TRUE, cex = 1.5)


cat("Simulated", simulationCorrelation(ar3), "\n", 
    "PACF value", pacfVal, "\n", 
    "Theoretical", 0.1, "\n")
```


##b)

```{r, warning=FALSE}
ar2 <- arima.sim(model = list(ar = c(0.8,0.1)), n = 100)

ywl <- ar(ar2, order = 2, method = "yule-walker", aic = FALSE)
cls <- ar(ar2, order = 2, method = "ols", aic = FALSE)
mle <- ar(ar2, order = 2, method = "mle", aic = FALSE)

df <- data.frame("Yule-Walker" = c(ywl$ar[1],ywl$ar[2]),
                 "Conditional LS" = c(cls$ar[1],cls$ar[2]),
                 "MLE" = c(mle$ar[1], mle$ar[2]))
rownames(df) <- c("Component 1", "Component 2")
df

mle <- arima(ar2, order = c(2,0,0), method = "ML")
ci <- unname(c(mle$coef[2]+1.96*(sqrt(mle$var.coef[2,2])),
        mle$coef[2]-1.96*(sqrt(mle$var.coef[2,2]))))
cat("Confidence Interval - ", ci, "\nPoint Estimate Value - ", 
    mle$coef[2], "\n", "Theoritical Value - ", 0.1, "\n")
print(ifelse(((0.1-ci[1])*(ci[2]-0.1)>0), 
       "Theoretical value fall within confidence",
       "Theoretical value does not fall within confidence"))

```

\newpage

##c)

```{r}
arimaSeason <- arima.sim(list(order = c(0,0,1),
                              ma = c(0,rep(0,10),0.6)), n = 200)
par(mfrow=c(1,2),oma = c(0, 0, 2, 0))
acf(arimaSeason, main=NA)
pacf(arimaSeason, main=NA)
mtext("Sample ACF and PACF", outer = TRUE, cex = 1.5)

#Theoretical
armaSeasonT <- ARMAacf(ma = c(0.3,rep(0,10),0.6,0.6*0.3))
armaSeasonTP <- ARMAacf(ma = c(0.3,rep(0,10),0.6,0.6*0.3), pacf = TRUE)

plot(armaSeasonT, type = "h", ylab = "ACF")
abline(h = 0, col = "blue") 
plot(armaSeasonTP, type = "h", ylab = "PACF") 
abline(h = 0, col = "blue")
mtext("Theoritical ACF and PACF", outer = TRUE, cex = 1.5)
```

\newpage

##d)

```{r}
par(mfrow=c(1,2),oma = c(0, 0, 2, 0))
seasonArimaSim = arima.sim(list(order = c(0,0,12), ma = c(0.7,rep(0,10),0.6)), n = 200)
seasonDF = data.frame(y = as.vector(seasonArimaSim), x = 1:200)

#Fitting arima model
seasonArimaFit = arima(seasonArimaSim, order = c(0,0,1), seasonal = list(order = c(0,0,1),period = 12))
seasonArimaPred = predict(seasonArimaFit, n.ahead = 30)

plot(seasonArimaSim, xlim = c(0,240), ylab = NA)
upper <- seasonArimaPred$pred + 1.96*seasonArimaPred$se
lower <- seasonArimaPred$pred - 1.96*seasonArimaPred$se
polygon(c(time(upper),rev(time(upper))),c(lower, rev(upper)),border = 8, col = "grey")
lines(seasonArimaPred$pred, col = "red")
abline(v = 200, col = "blue", lty = 1)

#Using gausspr to fit the data
seasonGaussModel = gausspr(y ~ x, data = seasonDF)
seasonGaussPred = predict(seasonGaussModel, newdata = data.frame(x = 1:230))


plot(seasonArimaSim, col = "black", type = "l",ylab = NA, xlab = "Time", xlim=c(0,240))
lines(seasonGaussPred, col = "red")
mtext("Arima Fit vs Gausspr on Seasonal data", outer = TRUE, cex = 1.5)
abline(v = 200, col = "blue", lty = 1)
```

\newpage

##e

```{r}
arimaE <- arima.sim(list(order = c(1,0,1), ar = 0.7, ma = 0.5), n = 50)

arimasample <- arimaE[1:40]

modelArima <- arima(arimasample, order = c(1,0,1), include.mean = 0)

pred101 <- predict(modelArima, n.ahead = 10)

ggplot() + 
  geom_line(aes(x=1:50, y=arimaE[1:50]), col="black", lwd=1) +
  geom_line(aes(x=41:50, y=arimaE[41:50]), col="blue", lwd=1) +
  geom_line(aes(x=41:50, y=pred101$pred[1:10]), col="darkgreen", lwd=1) +
  geom_line(aes(y = pred101$pred + 1.96*pred101$se, x = 41:50), col="red", lwd=1) +
  geom_line(aes(y = pred101$pred - 1.96*pred101$se, x = 41:50), col="red", lwd=1) + 
  geom_vline(xintercept = 40, col="gray", lwd=1) + 
  geom_hline(yintercept = 0, col="gray", lwd=1) + 
  ggtitle("Confidence Interval for Prediction on simulated data") +
  xlab("Time") + ylab("Value")
```

\newpage

#Question 2

```{r}
genPlots = function(x_t, dataset){
  par(mfrow=c(2,1),oma = c(0, 0, 2, 0))
  plot(x_t)
  plot(diff(x_t))
  mtext(paste(dataset, " dataset"), outer = TRUE, cex = 1.5)
  
  par(mfrow=c(2,2),oma = c(0, 0, 2, 0))
  acf(x_t, lag.max = 40, main="")
  pacf(x_t, lag.max = 40, main="")
  acf(diff(x_t), lag.max = 40, main="")
  pacf(diff(x_t), lag.max = 40, main="")
  mtext(paste(dataset, " dataset ACF and PACF plots"), outer = TRUE, cex = 1.5)
}
```

## Datasets

### Chicken

The decreasing trend in the ACF plot of chicken and the cutoff in the PCAF plot suggests that this could be an AR process. The positive ACF at lag 1 for the differenced data confirms that this is a AR process. The PCAF of the differenced data suggests that it is a AR(3) process, since there is a cutoff after lag 3 in the plot.
We can see seasonality in the differenced dataset. The arima model $ARIMA(3,1,0)(3,1,0)_{12}.$ would be good for this dataset.


```{r}
genPlots(chicken, "Chicken")
```

### so2

The decreasing trend in the ACF plot of the dataset suggests an ARIMA model could be a good fit for it. The negative ACF at lag 1 for the differenced dataset suggests to use an MA model. The pcaf plot of differenced dataset tells us that MA(7) model would be a good fit for the dataset. $ARIMA(0,1,7)$ would be a good model for this data.

```{r}
genPlots(so2, "so2")
```

### EQcount

The decreasing trend in the ACF plot of the dataset suggests an ARIMA model could be a good fit for it. The negative ACF at lag 1 for the differenced dataset suggests to use an MA model. The pcaf plot of differenced dataset tells us that MA(2) model would be a good fit for the dataset. $ARIMA(0,1,2)$ would be a good model for this data.

```{r}
genPlots(EQcount, "EQCount")
```

### HCT

The decreasing trend in the ACF plot of the dataset suggests an ARIMA model could be a good fit for it. The negative ACF at lag 1 for the differenced dataset suggests to use an MA model. The pcaf plot of differenced dataset tells us that MA(7) model would be a good fit for the dataset. $ARIMA(0,1,7)$ would be a good model for this data.
From the difference of order 1 we can see some seasonality after seven lags(seven days).

```{r}
genPlots(HCT, "HCT")
```